---
title: 'Assignment: Plot Deconstruction'
author: "Samantha Williams"
date: '2025-03-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Graphical Storytelling in R.
Exercise: Plot Deconstruction

# Part 1. Choose your data. 
Find a plot online, either in a scientific journal, a newspaper, or some other place (e.g. the world bank data portal) which:
Interests you and your partner.
You can improve on the plot. 

The data in the plot are complex enough for you to create an interesting graphic. I recommend at least two variables, perhaps more.
Post the URL of your chart to our subject Google Doc. Make sure no-one else has chosen the same plot as you.

# Part 2. Deconstruct the plot. 
Take a screenshot of the original plot
Think about the message of the plot, what were the authors trying to show?
Find at least two problems with the plot. Perhaps the visualisation is faulty (bar plot for continuous data, poor colour schemes, hidden outliers etc)
Write a caption for your plot that describes the message of the plot and the problems you found with it.
Download the data for the plot. To do this, you can either find their source data or use a manual data digitisation tool https://automeris.io/WebPlotDigitizer/ ) to build your CSV. This app allows you to upload a jpg of your plot, then follow the instructions to harvest data points from it.


```{r}
# install.packages("readr") - only do this once
library(readr) # to read csv files
library(ggplot2) # to graph
library(dplyr) # data manipulation
library(tidyr)  # for tidying
```

Load CSV files from https://www.sciencedirect.com/science/article/pii/S0022098103005136#FIG3

```{r}
bruvs <- read_csv("data/BRUVS.csv")
trawls <- read_csv("data/Trawls.csv")
```

Check the structure of the data and check X and Y values:
```{r}
head(bruvs)
head(trawls)
```

Rename the column names:
```{r}
colnames(bruvs) <- c("Transects", "NewSpecies")
colnames(trawls) <- c("Transects", "NewSpecies")
```

Mutate the graphs to add a column 'Method'
```{r}
bruvs <- bruvs %>% mutate(Method = "BRUVS")
trawls <- trawls %>% mutate(Method = "Trawls")
```

Combine the two data sets:
```{r}
data_combined <- bind_rows(bruvs, trawls) # because all column names are the same it stacks the data
```

Plot the data:
```{r}
ggplot() + 
  geom_point(data = data_combined, mapping = aes(x = Transects, y = NewSpecies)) + 
  geom_smooth(data = data_combined, mapping = aes(x = Transects, y = NewSpecies)) +
  theme_minimal() + 
  theme(
    axis.line = element_line(color = "black", size = 1),  # Add X and Y axis lines
    axis.text = element_text(size = 12),  # Make axis text bigger
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles
    legend.title = element_text(size = 12),  # Legend title styling
    legend.text = element_text(size = 10)  # Legend text styling
  ) +
```

```{r}
ggplot(data_combined, aes(x = Transects, y = NewSpecies, color = Method, fill = Method)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE, level = 0.99, linewidth = 1, alpha = 0.5) +
  labs(x = "Transects (N)", y = "New Species (N)") +
  theme_minimal() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 1),  
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    legend.background = element_blank()
  ) +
    scale_color_viridis_d(option = "cividis") +  
  scale_fill_viridis_d(option = "cividis") +  
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +  # More X-axis values
  scale_y_continuous(breaks = seq(30, 90, by = 10))  # More Y-axis values
```

Load colour blind safe colour scheme
```{r}
library(viridis)
```

# adding a fake set of error data 
Graph with grid lines and points
```{r}
ggplot(data_combined, aes(x = Transects, y = NewSpecies, color = Method, fill = Method)) +
  geom_point() +
  geom_ribbon(aes(ymin = NewSpecies - 3, ymax = NewSpecies + 3, fill = Method), alpha = 0.2, color = NA) +
  geom_smooth(method = "loess", se = TRUE, level = 0.8, linewidth = 1, alpha = 0.5) +
  labs(x = "Transects (N)", y = "New Species (N)", color = "Sampling Method", fill = "Sampling Method") +
  theme_minimal() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 1),  
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    legend.background = element_blank(),
  ) +
  scale_color_manual(values = c("BRUVS" = "darkblue", "Trawls" = "darkgreen")) +
  scale_fill_manual(values = c("BRUVS" = "darkblue", "Trawls" = "darkgreen")) +
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +  # More X-axis values
  scale_y_continuous(breaks = seq(30, 90, by = 10))  # More Y-axis values
```

Graph without gridlines or points

Change order of the legend so it matches graph:
```{r}
data_combined$Method <- factor(data_combined$Method, levels = c("Trawls", "BRUVS"))
```

```{r}
ggplot(data_combined, aes(x = Transects, y = NewSpecies, color = Method, fill = Method)) +
  geom_ribbon(aes(ymin = NewSpecies - 3, ymax = NewSpecies + 3, fill = Method), alpha = 0.2, color = NA) +
  geom_smooth(method = "loess", se = TRUE, level = 0.8, linewidth = 1, alpha = 0.5) +
  labs(x = "Transects (N)", y = "New Species (N)", color = "Sampling Method", fill = "Sampling Method") +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 1),  
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    legend.background = element_blank(),
  ) +
  scale_color_manual(values = c("Trawls" = "darkgreen", "BRUVS" = "darkblue")) +  
  scale_fill_manual(values = c("Trawls" = "darkgreen", "BRUVS" = "darkblue")) +
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +  # More X-axis values
  scale_y_continuous(breaks = seq(30, 90, by = 10))  # More Y-axis values
```


Try a different method to make error more realistic
```{r}
ggplot(data_combined, aes(x = Transects, y = NewSpecies, color = Method, fill = Method)) +
  geom_ribbon(aes(
    ymin = NewSpecies - (5 / sqrt(Transects)),  # Larger error at low Transects, smaller at high Transects
    ymax = NewSpecies + (5 / sqrt(Transects)),  
    fill = Method
  ), alpha = 0.2, color = NA) +
  geom_smooth(method = "loess", se = TRUE, level = 0.8, linewidth = 1, alpha = 0.5) +
  labs(x = "Transects (N)", y = "New Species (N)", color = "Sampling Method", fill = "Sampling Method") +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, size = 1),  
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    legend.background = element_blank(),
    legend.position = c(0.82, 0.22) # Puts the legened inside the graph
  ) +
  scale_color_manual(values = c("Trawls" = "darkgreen", "BRUVS" = "darkblue")) +  
  scale_fill_manual(values = c("Trawls" = "darkgreen", "BRUVS" = "darkblue")) +
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +  # More X-axis values
  scale_y_continuous(breaks = seq(30, 90, by = 10))  # More Y-axis values

```

# Part 3. Reconstruct a better version of the plot.
Fix the problems you described in your caption for the original plot. You will need to make an entirely new version of the plot using the skills you’ll learn in tomorrow's workshop. 

Don’t restrict yourself to the form of the original plot. Sometimes the data are not in the form of what you need for a better plot - we will learn in workshop 3 how to ‘tidy’ datasets.
Add more information to the plot to help convey the message more clearly. This could be additional data, reference markers or useful labels. Use the scientific literature for your guide (see some examples in the journal Nature)
Finally, write out a detailed caption for your new version of the plot. Describe it clearly, pointing out its main features and state the message to be gleaned from your plots. Include references or any other new data sources used in your plot.
Your e-portfolio will ultimately include the original plot and caption, and your new, better version of the plot immediately underneath. 

Original plot problems/caption:
We didn't like the error bars and found it hard to read. Also hard to distinguish between sampling methods. 


Species accumulation showing the number of new species detected with increasing transects for each sampling method (BRUVS and Trawls). Curves are based on repeated randomization of transect order, with error bars representing variability in species accumulation. However, the original plot had several issues: (1) The lack of color or line differentiation made it difficult to distinguish between methods, (2) The error bars appeared cluttered and inconsistent, making it challenging to interpret variability, and (3) The absence of a legend required direct labeling of lines, which is less clear in complex figures. 

# Caption: need to talk about adding a ribbon for confience interval because we didn't use the actual data as it was too hard to export. We identify this as a limitation of what we have done here. 
Species accumulation curves showing the number of new species detected with increasing transects for each sampling method (BRUVS and Trawls). Curves are based on repeated randomization of transect order, with shaded regions representing variability in species accumulation.
This revised version improves aspects by using colorblind-friendly colours, smoother confidence intervals, and a clearer legend for readability. Axis lines have also been tidied, by changing the theme to classic, making them bold and editing lablels. 

#Due date. This component of your e-portfolio should be ready to share in workshop 3 for this module, where we will have an interactive session to parade the plots to the class.  Your e-portfolio will ultimately include the original plot and caption, and your new, better version of the plot immediately underneath. 

