---
title: "Module 4 Assignment"
author: "Samantha Williams"
date: '2025-03-20'
output: html_document
---

### 3.10 Assignment: End-to-end data analysis in R 
For this report, real-world marine data collected as part of Queensland fisheries QFISH database was used.

#### Part 1. Obtaining Data
For this report, real-world marine data were sourced from the Queensland Government's [QFISH database](https://qfish.fisheries.qld.gov.au/), a publicly available resource that compiles commercial and recreational fisheries data. The dataset, sourced from the [Shark Control Program](https://qfish.fisheries.qld.gov.au/query/number-caught-by-area-calendar-year-and-species-group/table#), was used to explore patterns in species catch across areas and calendar years.

#### Part 2. Tidy the data
Before commencing the analysis, the raw dataset was transformed into a tidy format to ensure it was suitable for statistical exploration and visualization. The following steps outline the data wrangling process:

Install and load necessary packages:
```{r, eval = FALSE}
install.packages("tidyverse") # Rools for data cleaning, transformation, and plotting

install.packages("janitor") # Helpful functions for cleaning messy data (e.g., column names)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(viridis) # Colours for graphing
```


Load CSV file & begin to clean:
```{r message=FALSE, warning=FALSE}
sharkdata_raw <- 
  read_csv("../data/sharkdata_updated.csv") |>
  row_to_names(row_number = 2, remove_rows_above = FALSE) |> 
  clean_names()   # Standardise column names
```

Create new tables for each species (isolate them):
```{r}
sharkdata <- 
  sharkdata_raw |>
  select(area, contains("shark")) |> # select all everything with "shark" in 'area' row
  row_to_names(row_number = 1) |> # Use the first row of this subset as the new column names
  rename(Area = 1) |>  # Capitalise the first column 
  mutate(across(-c(Area), as.numeric)) # convert all to numeric except for 'Area'
  
turtledata <-
  sharkdata_raw |>
  select(area, contains("turtle")) |>
  row_to_names(row_number = 1) |>
  rename(Area = 1) |>
  mutate(across(-c(Area), as.numeric))

mammaldata <- 
  sharkdata_raw |>
  select(area, contains("mammal")) |>
  row_to_names(row_number = 1) |>
  rename(Area = 1) |>
  mutate(across(-c(Area), as.numeric))

otherdata <-
  sharkdata_raw |>
  select(area, contains("other")) |>
  row_to_names(row_number = 1) |>
  rename(Area = 1) |>
  mutate(across(-c(Area), as.numeric))
```

Convert into long format:
```{r}
sharkdata_long <- 
  sharkdata |>
  pivot_longer(!Area, names_to = "Year", values_to = "Count") |>
  mutate(Taxa = "Sharks") # Add a new column for Taxa for later processing

turtledata_long <- 
  turtledata |>
  pivot_longer(!Area, names_to = "Year", values_to = "Count") |>
  mutate(Taxa = "Turtles")

mammaldata_long <-
  mammaldata |>
  pivot_longer(!Area, names_to = "Year", values_to = "Count") |>
  mutate(Taxa = "Mammals")

otherdata_long <-
  otherdata |>
  pivot_longer(!Area, names_to = "Year", values_to = "Count") |>
  mutate(Taxa = "Other")
```

#### Part 3. Basic Statistics 
Summarise data by computing mean and standard deviation, and add a Taxa identifier:
```{r}
shark_stats <- 
  sharkdata_long |>
  filter(Area != "Grand Total") |> # Removes unecessary row
  group_by(Year) |> # Organise data by each unique year
  summarise(Mean = mean(Count, na.rm = TRUE), # Calculate Mean
            SD = sd(Count, na.rm = TRUE)) |> # Calculate Standard Deviation from the Mean
  mutate(Year = as.numeric(Year)) |>
  mutate(Taxa = "Sharks") # Adds a column for taxa 

turtle_stats <- 
  turtledata_long |>
  filter(Area != "Grand Total") |>
  group_by(Year) |>
  summarise(Mean = mean(Count, na.rm = TRUE),
            SD = sd(Count, na.rm = TRUE)) |>
  mutate(Year = as.numeric(Year)) |>
  mutate(Taxa = "Turtles")

mammal_stats <- 
  mammaldata_long |>
  filter(Area != "Grand Total") |>
  group_by(Year) |>
  summarise(Mean = mean(Count, na.rm = TRUE),
            SD = sd(Count, na.rm = TRUE)) |>
  mutate(Year = as.numeric(Year)) |>
  mutate(Taxa = "Mammals")

other_stats <-
  otherdata_long |>
  filter(Area != "Grand Total") |>
  group_by(Year) |>
  summarise(Mean = mean(Count, na.rm = TRUE),
            SD = sd(Count, na.rm = TRUE)) |>
  mutate(Year = as.numeric(Year)) |>
  mutate(Taxa = "Other")
```

Combine statistics into one table:
```{r}
combined_stats <- 
  bind_rows(shark_stats, turtle_stats, mammal_stats, other_stats)
```

Combined taxa count data:
```{r}
combined_data <-
  bind_rows(sharkdata_long, turtledata_long, mammaldata_long, otherdata_long) |>
  mutate(Year = as.integer(Year))
```

#### Part 4. Plotting Data

Create plot for mean species caught:
```{r annual-catch-plot, message=FALSE, warning=FALSE}
annual_catch_plot <- ggplot(data = combined_stats, aes(x = Year, y = Mean, color = Taxa)) +
  geom_pointrange(aes(ymin = Mean - SD, ymax = Mean + SD),
                  position = position_dodge(width = 0.5)) +
  labs(y = "Mean Annual Catch", color = "Taxa") +
  theme_classic() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    panel.border = element_rect(color = "black", fill = NA, size = 1)
  ) +
  scale_color_viridis_d(option = "D")
```


Create plot for species caught at each location:
```{r taxa-trends-plot, message=FALSE, warning=FALSE}
taxa_trends_plot <- ggplot(combined_data, aes(x = Year, y = Count, color = Area)) +
  geom_line() +
  labs(
    color = "Area",
    y = "Taxa Caught") +
  facet_wrap(~ Taxa, ncol = 2) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.title = element_text(size = 13, face = "bold"),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 9)
  )
```


#### Part 5. Outputs
Could not upload a markdown document for figures so for reference this is the code to add figure captions and dimensions:
Figure 1.
```{r}
# ```{r annual-catch-output, fig.cap="Figure 1. Mean annual catch (± SD) of marine taxa recorded by the Queensland Shark Control Program from 2001 to 2025. Each point represents the average number of individuals caught per year across all regions for a given taxonomic group (Mammals, Other, Sharks, and Turtles). Shark catch rates were consistently higher than those of other groups throughout the time series.", fig.height=120/25.4, fig.width=170/25.4, message=FALSE, warning=FALSE}
#annual_catch_plot 
# ```
```

Figure 2.
```{r}
#```{r taxa-trends-output, fig.cap="Figure 2. Trends in taxa caught across Queensland regions from 2001 to 2025. The data are faceted by taxonomic group (Mammals, Other, Sharks, Turtles), with coloured lines representing different catch areas. Shark catches are consistently higher than other groups, particularly in Mackay.", fig.height=120/25.4, fig.width=170/25.4, message=FALSE, warning=FALSE}
#taxa_trends_plot
#```
```

Export figures:
```{r message=FALSE, warning=FALSE}
ggsave("annual_catch_plot.png", plot = annual_catch_plot, device = "png", width = 240, height = 160, units = "mm", dpi = 300)
```

```{r message=FALSE, warning=FALSE}
ggsave("taxa_trends_plot.png", plot = taxa_trends_plot, device = "png", width = 240, height = 160, units = "mm", dpi = 300)
```



