---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Samantha Williams"
output: html_document
date: "2025-03-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Install and load tidyverse packages
```{r}
## install.packages("tidyverse") # Delete this line once installed
#library("tidyverse")
```


# Load the data (mpg is built into ggplot2)
```{r}
data(mpg) 
```


# Quick data checks
```{r}
head(mpg) # creates a tibble
glimpse(mpg) # see every column in the data frame
summary(mpg) # provides a function summary 
```


# Create first plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) # shows a very basic plot for engiine size and fuel efficiency.
```

The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?
As engine size increases, fuel efficiency decreases. 


# Understanding grammar of graphics
What happens if we just try to run ggplot on its own?
```{r}
ggplot() 
# Or with just the data specified?
ggplot(data = mpg) # when only ggplot is run on it's own it doesn't do anything, as it hasn't been told to do anything with the data. 
```

We need to map some aesthetics!!

# When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics.
# Change point colour by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) # creats a scatter plot with colours for each class.
```

# Change point size by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class)) # creates a warning message: using size for a discrete variable is not advised. 
```
Note the warning!!!


# Change transparency (alpha) by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) # changes gradient of points with no colour
# warning message: using alpha for a discrete variable is not advised. 
```
Another warning!!
Question: When would using alpha (or size be appropriate??)
It would be appropriate when using continuous values or to see overlapping values so you can see more. 

# Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class)) # creates a really messy graph
```


# Make all points blue
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") #all points are blue
```


# Troubleshooting 

```{r}
#ggplot(data = mpg) 
#+ geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) # correct code
# the + should be on top line
```

# Faceting
A common practice is to break a single complex plot into many sub plots to develop separate plots for a range of reasons, i.e., to show a subset of data.
## facet_wrap
Split the plot out by car type (class)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2) # creates 7 basic smaller graphs sorted by class
```


## facet_grid
A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl) # confusing to read 
```
Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!


## facet_grid by just row (or column)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(cyl ~ .)
  #facet_grid(. ~ cyl) # Alternatively

```

# Exercise:
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 
nrow = number of rows
ncol = number of columns 
switch = changes where labels are displayed on the top

# Lines

We already used points:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) # basic point plot
```


However, ggplot2 can use a variety of geom objects to represent the data. Here, we might want to use bar plots, line charts, boxplots and so on. Well we can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 

# To display the same data as a smooth line fit through the points use geom_smooth().
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # creates a smooth plot 
```
### Question: whey don't we use geom_line() here? What would that look like? ###
A geom_line would create a line between each data point as follows:
```{r}
ggplot(data = mpg) + 
  geom_line(mapping = aes(x = displ, y = hwy)) # this is very messy and hard to read.
```

So let’s recap. A geom is an object that your plot uses to represent the data. To change the geom type in your plot, simply change the geom function that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 

## Using comments (#)
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

Question: how does R work out where to plot the line??? Can use the chunk output to inform us. Can also use the help menu.
Calculation is performed by predicting values based off the data we have provided - by the predictdf() generic and its methods. 

# Changing linetype
Creates a plot with different types of lines, e.g., dashed, etc. for different subsets of data. 
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) # Can also use "lty = " for the same result

```

# Grouping
If you want to group objects by a categorical variable (like species or sex or site), you can use the the group argument to show that our data is grouped. It won’t add a new legend but can be used to vary plots by the discrete variable.
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv)) # plots different groups without different line types and no legend
```
 
# Change line colour based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  ) # same line type but the colour of each line changes
```

# Multiple geoms
We already did this one way
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
# creates a scatter plot and a smooth line plot running through the same plot. 
```

A better way...
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
# much more clean and easier to read and therefore troubleshoot. 
```


# Can still manipulate each geom/layer separately:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
# adds different coloured plots for each class type
```

Now we have different colours to indicate the raw data (geom_point) and one colour for the overall relationship (geom_smooth). Neat!

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```
Question: what does the "se = FALSE" do?
It removed the shaded region around the smooth line. It likely stands for standard error (the confidence interval) and false tells it not to display it.

# Exercise:
1.  What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
geom_line() for line chart 
geom_boxplot() for box plot
geom_histogram() for histogram
geom_area() for area

2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
```{r}
ggplot(data = mpg, mapping = aes(y = hwy)) +
  geom_histogram()
#creates a very basic histogram

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_boxplot() #basic box plot with one x variable 

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_area() # creates an area chart - not great for visualisation
```


3.  Will these two graphs look different? Why/why not?
No because you're just doing the same thing in the second chunk of code, but just the long version twice. By establishing the aesthetic early on it applies it to the entire chunk of code. But the second option allows you to make the two different geoms have different features. The first option is much more clean and tidy for what we want here. 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

# Transformations and stats
The diamonds dataset comes in ggplot2 and contains information about ~54,000 diamonds, including the price, carat, color, clarity, and cut of each diamond.
## Load data
```{r}
data("diamonds")
glimpse(diamonds) # shows rows and columns in dataset
```

# Bar plot
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut)) #maps bar plot showing count of each cut. 
```

# Overriding defaults
Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo # creates and shows tibble made. 
```

Map the new data we just made:
```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
Question: Why did we specify now specify a y axis?
Because we created a new variable with specific counts. To generate a geom_bar() we need a y aesthetic. The existing dataset didn't need y to be defined because R just counted the occurrences of each cut, but our created datset only had one column for counts and therefore couldn't just count how often each occurred. 

# What if we want to know the proportion rather than the absolute frequency?
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1)) # no error with this code 
```

Question: does anyone get the warning that's in the workbook? Yes.
Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
If so, can use # geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1))

# Plotting statistical details
Stat_summary() can compute any aggregate. 
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


# Aesthetic adjustments adjustments
Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colours.

Using colour = __ highlights each bar with a different colour by cut. 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
```

Fill makes the whole bar in the plot a different colour (by cut).
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

Question: Does anyone notice anything different in the colour scheme of this plot?
(Hint: It's in the viridis colour palette (colour blind friendly), but why is it different from the colour palette we used earlier?)
Check out the difference:
This plot describes different variables entirely, so it makes sense to use completely different colours to show the reader that the different colours represent different things. 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 
```

Whereas, this plot describes continuous values - so it makes more sense to have a continuous colour scheme, i.e., variations of the same colour like greens, yellows and blues to make more of a gradient. 
```{r}
ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```


# Filling by a variable
Now try using these aesthetics to colour by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a position argument.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity)) # creates different bars in a plot with gradients within each bar for levels of clarity. 
```


# To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
# makes the gradients of colour within each bar transparent. 
```


# To color the bar outlines with no fill color
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
# also hard to read, as the boarders of each bar are quite thin and background had a grid. 
```


# Position adjustments
position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

# position = "dodge" 
Places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
# creates a lot more blank space. 
```

# Jittering
position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. One bonus position adjustment is jitter, which slightly moves points so you can see them all. This is useful for scatterplots but not barplots. 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter") +
  geom_point(mapping = aes(x = displ, y = hwy))
# almost makes it look fuzzy. 
```










